### 2014-12-09 doc/thesis/typedlua.tex
Improvements to the text about table types
Minor improvements to chapters 2 and 3

\section{Optional type annotations}

@@

\begin{figure}[!ht]
\textbf{Types}\\
\dstart
\begin{align*}
\textit{type} & ::= \; \textit{primarytype} \; [\texttt{`?'}]\\
\textit{primarytype} & ::= \; \textit{literaltype} \; | \;
  \textit{basetype} \; | \;
  \textbf{nil} \; | \;
  \textbf{value} \; | \;
  \textbf{any} \; | \;
  \textbf{self} \; | \;
  \textit{Name}\\
& | \; \textit{functiontype} \; | \;
  \textit{tabletype} \; | \;
  \textit{primarytype} \; \texttt{`|'} \; \textit{primarytype}\\
\textit{literaltype} & ::= \; \textbf{false} \; | \;
  \textbf{true} \; | \;
  \textit{Int} \; | \;
  \textit{Float} \; | \;
  \textit{String}\\
\textit{basetype} & ::= \; \textbf{boolean} \; | \;
  \textbf{integer} \; | \;
  \textbf{number} \; | \;
  \textbf{string}\\
\textit{functiontype} & ::= \; \textit{tupletype} \; \texttt{`->'} \; \textit{rettupletype}\\
\textit{methodtype} & ::= \; \textit{tupletype} \; \texttt{`=>'} \; \textit{rettupletype}\\
\textit{tabletype} & ::= \; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
\textit{tupletype} & ::= \; \texttt{`('} \; [typelist] \; \texttt{`)'}\\
\textit{typelist} & ::= \; \textit{type} \; \{\texttt{`,'} \; \textit{type}\} \; [\texttt{`*'}]\\
\textit{rettupletype} & ::= \; \textit{unionlist} \; [\texttt{`?'}]\\
\textit{unionlist} & ::= \; \textit{tupletype} \; | \;
  \textit{unionlist} \; \texttt{`|'} \; \textit{unionlist}\\
\textit{tabletypebody} & ::= \; \textit{type} \; | \; \textit{fieldtypelist}\\
\textit{fieldtypelist} & ::= \; [\textbf{const}] \; \textit{fieldtype} \; \{\texttt{`,'} \; [\textbf{const}] \; \textit{fieldtype}\}\\ 
\textit{fieldtype} & ::= \; \textit{keytype} \; \texttt{`:'} \; \textit{type}\\
\textit{keytype} & ::= \; \textit{literaltype} \; | \;
  \textit{basetype} \; | \;
  \textbf{value} \; | \;
  \textbf{any}\\
\textit{rettype} & ::= \; \textit{type} \; | \; \textit{rettupletype}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua types}
\label{fig:types}
\end{figure}

Figure \ref{fig:types} presents the concrete syntax of the
Typed Lua types in extended BNF.
We classify Typed Lua types into two categories:
\emph{first-level types} and \emph{second-level types}.
First-level types consist of \emph{type} and represent Lua values,
while second-level types consist of either \emph{tupletype} or
\emph{rettype} and represent the type of expression lists,
multiple assignments, and function applications.
First-level types include literal types, base types, the type \texttt{nil},
the top type \texttt{value}, the dynamic type \texttt{any},
the self type \texttt{self}, named types, function types, table types,
and union types.
Second-level types include the type void \texttt{()},
vararg types, tuple types, and union of tuple types.

@@

In this example, Typed Lua cannot guarantee that \texttt{distance} is never
going to call \texttt{absolute} with a parameter that is not an integer number,
because in the semantics of Lua the minus operator can result in a
value that is not an integer number.
More precisely, we can call \texttt{absolute} inside \texttt{distance}
because the expression \texttt{x - y} has type \texttt{any},
and it is consistent with type \texttt{integer}.
Still, the dynamic type \texttt{any} may be hiding a value of a type
that is not an integer, making the dynamically typed code crash the
statically typed code.
This is a typical example where run-time checks would ensure safety
between the interaction of dynamically typed and statically typed code.

\section{Functions}

@@

As we can see in Figure \ref{fig:types}, second-level types consist
of either \emph{tupletype} or \emph{rettupletype}.
A second-level type is the type void \texttt{()},
a tuple of first-level types optionally ending in a variadic type,
or a union of these tuples.
A variadic type \texttt{t*} is a generator for a sequence of values
of the union type \texttt{t|nil}.
Union of tuples may appear only on the return type of functions,
and together with \emph{projection types} play an important role
in functions that are overloaded on the return type.
We will explain union types and projection types in more detail
in the next section.

@@

In the default mode, the type of \texttt{multiple} is
\texttt{(value*) -> (integer, string, nil*)},
and the type of \texttt{sum} is \texttt{(integer, integer, value*) -> (integer, nil*)}.
In the right hand side of the multiple assignment, only the first
value produced by the first call to multiple gets used,
so the type of the right hand side is
\texttt{(integer, integer, string, nil*)},
and the type assigned to \texttt{x}, \texttt{y}, and \texttt{z}
are respectively \texttt{integer}, \texttt{integer}, and \texttt{string}.
The same tuple is assigned to the call to \texttt{sum} and it
compiles without errors, as the first two components of the tuple
are consistent with the types of the parameters, and the other
components are consistent with \texttt{value}.

@@

\section{Unions}

@@

Lua programmers also overload the return type of functions,
usually for signaling the occurrence of errors.
In this idiom, a function returns its normal set of return values
when it successfully finished its execution,
and usually returns \texttt{nil} as the first value plus an
error message or other data that describes the error
when something failed during its execution.
Next, we show an example:
\begin{verbatim}
    local function idiv (dividend:integer, divisor:integer):
          (integer, integer)|(nil, string)
      if divisor == 0 then
        return nil, "division by zero"
      else
        local r = dividend % divisor
        local q = (dividend - r) / divisor
        return q, r
      end 
    end
\end{verbatim}

@@

Projection types do not appear in the subtyping and consistent-subtyping
relations because Typed Lua actually replaces them with the union of the
corresponding component of each tuple.
If a variable with a projection type appears in a \texttt{type} predicate,
it discriminates against all tuples in the union.
In our example, $X$ is \texttt{(integer, integer, nil*)|(nil, string, nil*)}
outside of the \texttt{if} statement,
but \texttt{(integer, integer, nil*)} inside the \texttt{then} block,
and \texttt{(nil, string, nil*)} inside the \texttt{else} block.
Thus, variable \texttt{q} has type \texttt{integer} and variable \texttt{r} also
has type \texttt{integer} inside the \texttt{then} block;
and variable \texttt{q} has type \texttt{nil} while variable \texttt{r}
has type \texttt{string} inside the \texttt{else} block.

We could also have used \texttt{type(q) == "number"} or
even \texttt{type(r) == "number"} as the predicate of our example,
as both predicates would produce the same result.
However, the form that appears in our example is much more succinct and idiomatic.

@@

The overloading mechanism of Typed Lua have a limitation:
the return type cannot depend on the input type.
Next, we show an example:
\begin{verbatim}
    local function limitation (x:number|string, y:number|string)
      if type(x) == "number" then
        if type(y) == "number" then
          return x + y
        end
      else
        if type(y) == "string" then
          return x .. y
        end
      end
    end
\end{verbatim}

This example shows that we cannot write a function that is
guaranteed to return a number when we pass two numbers and
guaranteed to return a string when we pass two strings.
Even though we could use intersection types to express the type
of this function as
\texttt{(number, number) -> (number) \string&
(string, string) -> (string)},
and therefore solve this limitation,
intersection types require more sophisticated flow typing to
check whether a function has this type,
and we still need to work on this problem.

\section{Tables and Interfaces}

@@

\section{Modules}

@@

After we define the module \texttt{mymath}, regardless of adopted style,
users can use it in the standard way:
\begin{verbatim}
    local m = require "mymath"
    print(m.pow(2, 3))                   -- 8
    print(m.pow(2, "foo"))               -- compile-time error
\end{verbatim}

And now the users of the module \texttt{mymath} have the policies that
Typed Lua provides.
The call to \texttt{require} assigns the type of the module \texttt{mymath}
to \texttt{m}, so Typed Lua can catch misuses of the module.

@@

\section{Object-Oriented Programming}

@@

Typed Lua uses table types along with the self type to
represent objects, as we show in the following example:
\begin{verbatim}
    local Shape = { x = 0.0, y = 0.0 }
    const function Shape:move (dx:number, dy:number)
      self.x = self.x + dx
      self.y = self.y + dy
    end
\end{verbatim}

@@

In the same example, we also declare the interface \texttt{Shape}.
Notice that we use double arrow instead of single arrow inside the
definition of the interface \texttt{Shape}.
The double arrow is syntactic sugar for defining the type of methods,
as it includes an implicit first parameter of type \texttt{self}.
The interface definition allows us to export the type of the class \texttt{Shape},
which is the same type that is given by the variable \texttt{Shape}:
\begin{verbatim}
    { "x":number, "y":number,
      const "new":(self, number, number) -> (self),
      const "move":(self, number, number) -> () }
\end{verbatim}

@@

\section{Description files}

@@

\begin{figure}[!ht]
\textbf{The complete syntax of description files}\\
\dstart
\begin{align*}
\textit{description} & ::= \; \textit{desclist}\\
\textit{desclist} & ::= \; \textit{descitem} \; \{ \textit{descitem} \}\\
\textit{descitem} & ::= \; \textit{typedid} \; | \;
  \textit{interface} \; | \;
  \textit{userdata}\\
\textit{typedid} & ::= \; \textit{id} \; \texttt{`:'} \; \textit{type}\\ 
\textit{interface} & ::= \; \textbf{interface} \; \textit{typedec}\\
\textit{userdata} & ::= \; \textbf{userdata} \; \textit{typedec}\\
\textit{typedec} & ::= \; \textit{Name} \; { \; \textit{decitem} \; } \; \textbf{end}\\
\textit{decitem} & ::= \; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
\textit{idtype} & ::= \; \textit{type} \; | \; \textit{methodtype}\\
\textit{idlist} & ::= \; \textit{id} \; \{ \; \texttt{`,'} \; \textit{id} \; \}\\
\textit{id} & ::= \; [\textbf{const}] \; \textit{Name}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua description files}
\label{fig:tldsyntax}
\end{figure}

Figure \ref{fig:tldsyntax} shows the complete syntax of Typed Lua
description files in extended BNF.
A Typed Lua description file defines the table type that represents
a certain module and the type names that are exported
along with this table type.
We can export a type name through the declaration of an interface or
userdata.
As we saw, an interface declaration is just a shortcut to a table type.
An userdata declaration is a shortcut to a table type that includes
a \emph{brand} that is its name.
The Typed Lua compiler uses this brand to combine structural with
nominal type checking, so two different userdata are not subtype
of each other because they do not share the same brand.
The Typed Lua compiler searches for a description file when it cannot
not find the respective Typed Lua file that is the argument of
\texttt{require}.

The following example shows the description file for \texttt{lmd5} \citep{lmd5},
a MD5 digest library for Lua that is written in C:
\begin{verbatim}
    userdata md5_context
      __tostring : (self) -> (string)
      clone : (self) -> (self)
      digest : (self, value) -> (string)
      new : (self) -> (self)
      reset : (self) -> (self)
      update : (self, string*) -> (self)
    end

@@

This description file exports the type \texttt{md5\string_context}
through the \texttt{userdata} declaration, and the table type
that represents the module.
Now we can use the Typed Lua compiler to check for type errors on
our use of the library:
\begin{verbatim}
    local m = require "md5"
    local x = m.new()
    local y = x:clone()
    local z = m.clone("foo")             -- compile-time error
    print(x:digest() == m.digest(x))
\end{verbatim}

In this example there is a type error because function \texttt{clone}
expects a value of type \texttt{md5\string_context} instead of a value
of type \texttt{string}.

The description files are the mechanism that we used to include the
typing of the Lua standard libraries inside Typed Lua.

>>>>

\section{Optional type annotations}
\label{sec:optann}

@@

\begin{figure}[!ht]
\textbf{Types}\\
\dstart
\begin{align*}
\textit{type} ::= & \;\; \textit{primarytype} \; [\texttt{`?'}]\\
\textit{primarytype} ::= & \;\; \textit{literaltype} \; | \;
  \textit{basetype} \; | \;
  \textbf{nil} \; | \;
  \textbf{value} \; | \;
  \textbf{any} \; | \;
  \textbf{self} \; | \;
  \textit{Name}\\
& | \; \textit{functiontype} \; | \;
  \textit{tabletype} \; | \;
  \textit{primarytype} \; \texttt{`|'} \; \textit{primarytype}\\
\textit{literaltype} ::= & \;\; \textbf{false} \; | \;
  \textbf{true} \; | \;
  \textit{Int} \; | \;
  \textit{Float} \; | \;
  \textit{String}\\
\textit{basetype} ::= & \;\; \textbf{boolean} \; | \;
  \textbf{integer} \; | \;
  \textbf{number} \; | \;
  \textbf{string}\\
\textit{functiontype} ::= & \;\; \textit{tupletype} \; \texttt{`->'} \; \textit{rettype}\\
\textit{tupletype} ::= & \;\; \texttt{`('} \; [typelist] \; \texttt{`)'}\\
\textit{typelist} ::= & \;\; \textit{type} \; \{\texttt{`,'} \; \textit{type}\} \; [\texttt{`*'}]\\
\textit{rettype} ::= & \;\; \textit{type} \; | \;
  \textit{uniontuple} \; [\texttt{`?'}]\\
\textit{uniontuple} ::= & \;\; \textit{tupletype} \; | \;
  \textit{uniontuple} \; \texttt{`|'} \; \textit{uniontuple}\\
\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
\textit{tabletypebody} ::= & \;\; [\textit{keytype} \; \texttt{`:'}] \; \textit{type} \; | \;
  \textit{recordtype}\\
\textit{keytype} ::= & \;\; \textit{basetype} \; | \;
  \textbf{any}\\
\textit{recordtype} ::= & \;\; \textit{recordfield} \; \{\texttt{`,'} \; \textit{recordfield}\} \; [\texttt{`,'} \; \textit{type}]\\
\textit{recordfield} ::= & \;\; [\textbf{const}] \; \textit{literaltype} \; \texttt{`:'} \; \textit{type}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua types}
\label{fig:types}
\end{figure}

Figure \ref{fig:types} presents the concrete syntax of the
Typed Lua types in extended BNF.
We classify Typed Lua types into two categories:
\emph{first-level types} and \emph{second-level types}.
First-level types consist of \emph{type} and represent Lua values,
while second-level types consist of either \emph{tupletype} or
\emph{rettype} and represent the type of expression lists,
multiple assignments, and function applications.
First-level types include literal types, base types, the type \texttt{nil},
the top type \texttt{value}, the dynamic type \texttt{any},
the self type \texttt{self}, named types, function types, table types,
and union types.
Second-level types include the type void \texttt{()},
vararg types, tuple types, and unions of tuple types.

@@

In this example, Typed Lua cannot guarantee that \texttt{distance} is never
going to call \texttt{absolute} with a parameter that is not an integer number,
because in the semantics of Lua the minus operator can result in a
value that is not an integer number.
In fact, we can overload the minus operator to return a value that is
not even a number.
However, we can call \texttt{absolute} inside \texttt{distance}
because the expression \texttt{x - y} has type \texttt{any},
and it is consistent with type \texttt{integer}.
Still, the dynamic type \texttt{any} may be hiding a value of a type
that is not an integer, making the dynamically typed code crash the
statically typed code.
This is a typical example where run-time checks would ensure safety
between the interaction of dynamically typed and statically typed code.

\section{Functions}
\label{sec:func}

@@

As we can see in Figure \ref{fig:types}, second-level types consist
of either \emph{tupletype} or \emph{rettype}.
A second-level type is the type void \texttt{()},
a tuple of first-level types optionally ending in a variadic type,
or a union of these tuples.
A variadic type \texttt{t*} is a generator for a sequence of values
of the union type \texttt{t|nil}.
Union of tuples may appear only on the return type of functions,
and together with \emph{projection types} play an important role
in functions that are overloaded on the return type.
We will explain union types and projection types in more detail
in the next section.

@@

In the default mode, the type of \texttt{multiple} is
\texttt{(value*) -> (integer, string, nil*)},
and the type of \texttt{sum} is \texttt{(integer, integer, value*) -> (integer, nil*)}.
In the right hand side of the multiple assignment, only the first
value produced by the first call to multiple gets used,
so the type of the right hand side is
\texttt{(integer, integer, string, nil*)},
and the types assigned to \texttt{x}, \texttt{y}, and \texttt{z}
are \texttt{integer}, \texttt{integer}, and \texttt{string}, respectively.
The same tuple is assigned to the call to \texttt{sum} and it
compiles without errors, as the first two components of the tuple
are consistent with the types of the parameters, and the other
components are consistent with \texttt{value}.

@@

\section{Unions}
\label{sec:unions}

@@

Lua programmers also overload the return type of functions,
usually for signaling the occurrence of errors.
In this idiom, a function returns its normal set of return values
when it successfully finished its execution,
and usually returns \texttt{nil} as the first value plus an
error message or other data that describes the error
when something failed during its execution.
Next, we show an example:
\begin{verbatim}
    local function idiv (dividend:integer, divisor:integer):
          (integer, integer)|(nil, string)
      if divisor == 0 then
        return nil, "division by zero"
      else
        local r = dividend % divisor
        local q = (dividend - r) // divisor
        return q, r
      end 
    end
\end{verbatim}

@@

Typed Lua replaces projection types with the union of the corresponding
component of each tuple when it infers that a variable has a projection type.
If a variable with a projection type appears in a \texttt{type} predicate,
it discriminates against all tuples in the union.
In our example, $X$ is \texttt{(integer, integer, nil*)|(nil, string, nil*)}
outside of the \texttt{if} statement,
but \texttt{(integer, integer, nil*)} inside the \texttt{then} block,
and \texttt{(nil, string, nil*)} inside the \texttt{else} block.
Thus, variable \texttt{q} has type \texttt{integer} and variable \texttt{r} also
has type \texttt{integer} inside the \texttt{then} block;
and variable \texttt{q} has type \texttt{nil} while variable \texttt{r}
has type \texttt{string} inside the \texttt{else} block.

We could also have used \texttt{type(q) == "integer"} or
even \texttt{type(r) == "integer"} as the predicate of our example,
as both predicates would produce the same result.
However, the form that appears in our example is much more succinct and idiomatic.
Note that the type \texttt{integer} is restricted to Lua 5.3,
as we use the function \texttt{math.type} to decide whether a number
is integer or not.

@@

The overloading mechanism of Typed Lua have a limitation:
the return type cannot depend on the input type.
Next, we show an example:
\begin{verbatim}
    local function limitation (x:number|string)
      if type(x) == "number" then
        return x + x
      else
        return x .. x
      end
    end
\end{verbatim}

This example shows that we cannot write a function that is
guaranteed to return a number when we pass a number and
guaranteed to return a string when we pass a string.
Even though we could use intersection types to express the type
of this function as
\texttt{(number) -> (number) \string&
(string) -> (string)},
and therefore solve this limitation,
intersection types require more sophisticated flow typing to
check whether a function has this type,
and we still need to work on this problem.

\section{Tables and Interfaces}
\label{sec:tables}

@@

\section{Modules}
\label{sec:modules}

@@

After we define the module \texttt{mymath}, regardless of adopted style,
users can use it in the standard way:
\begin{verbatim}
    local mymath = require "mymath"
    print(mymath.pow(2, 3))              -- 8
    print(mymath.pow(2, "foo"))          -- compile-time error
\end{verbatim}

And now the users of the module \texttt{mymath} have the policies that
Typed Lua provides.
The call to \texttt{require} assigns the type of the module \texttt{mymath}
to the local \texttt{mymath}, so Typed Lua can catch misuses of the module.

@@

\section{Object-Oriented Programming}
\label{sec:oop}

@@

Typed Lua uses table types along with the \texttt{self} type to
represent objects, as we show in the following example:
\begin{verbatim}
    local Shape = { x = 0.0, y = 0.0 }
    const function Shape:move (dx:number, dy:number)
      self.x = self.x + dx
      self.y = self.y + dy
    end
\end{verbatim}

@@

In the same example, we also declare the interface \texttt{Shape}.
Notice that we use double arrow instead of single arrow inside the
definition of the interface \texttt{Shape}.
The double arrow is syntactic sugar for defining the type of methods,
as it includes an implicit first parameter of type \texttt{self}.
The double arrow can appear only inside the declaration of
interfaces and userdata, which we will introduce in the next section.
The interface definition allows us to export the type of the class \texttt{Shape},
which is the same type that is given by the variable \texttt{Shape}:
\begin{verbatim}
    { "x":number, "y":number,
      const "new":(self, number, number) -> (self),
      const "move":(self, number, number) -> () }
\end{verbatim}

@@

\section{Description files}
\label{sec:tld}

@@

\begin{figure}[!ht]
\textbf{The complete syntax of description files}\\
\dstart
\begin{align*}
\textit{description} ::= & \; \textit{desclist}\\
\textit{desclist} ::= & \; \textit{descitem} \; \{ \textit{descitem} \}\\
\textit{descitem} ::= & \; \textit{typedid} \; | \;
  \textit{interface} \; | \;
  \textit{userdata} \; | \;
  \textit{typealias}\\
\textit{typedid} ::= & \; \textit{id} \; \texttt{`:'} \; \textit{type}\\ 
\textit{interface} ::= & \; \textbf{interface} \; \textit{typedec}\\
\textit{userdata} ::= & \; \textbf{userdata} \; \textit{typedec}\\
\textit{typedec} ::= & \; \textit{Name} \; { \; \textit{decitem} \; } \; \textbf{end}\\
\textit{decitem} ::= & \; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
\textit{idtype} ::= & \; \textit{type} \; | \; \textit{methodtype}\\
\textit{idlist} ::= & \; \textit{id} \; \{ \; \texttt{`,'} \; \textit{id} \; \}\\
\textit{id} ::= & \; [\textbf{const}] \; \textit{Name}\\
\textit{typealias} ::= & \; \textbf{typealias} \; \textit{Name} \; \texttt{`='} \; \textit{type}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua description files}
\label{fig:tldsyntax}
\end{figure}

Figure \ref{fig:tldsyntax} shows the complete syntax of Typed Lua
description files in extended BNF.
A Typed Lua description file defines the table type that represents
a certain module and the type names that are exported
along with this table type.
We can export a type name through the declaration of an interface,
userdata, or type alias.
As we saw, an interface declaration is just a shortcut to a table type
that represents a record.
An userdata declaration is similar to an interface declaration,
but it also includes its name as the \emph{brand} of the table type.
The Typed Lua compiler uses this brand to combine structural with
nominal type checking, so two different userdata are not subtype
of each other because they do not share the same brand.
A type alias declaration assigns a name to a first-level type.
Even though the type alias declaration may make the interface declaration
look redundant, it does not have a syntax as nice as the syntax
of the interface declaration for defining records.
The type alias declaration is specially useful when we need to create
names for types that are not records.

The Typed Lua compiler searches for a description file when it cannot
not find the respective Typed Lua file that is the argument of
\texttt{require}.

The following example shows the description file for \texttt{lmd5} \citep{lmd5},
a MD5 digest library for Lua that is written in C:
\begin{verbatim}
    userdata md5_context
      __tostring : (self) -> (string)
      clone : (self) -> (self)
      digest : (self, value) -> (string)
      new : (self) -> (self)
      reset : (self) -> (self)
      update : (self, string*) -> (self)
      version : string
    end

@@

This description file exports the type \texttt{md5\string_context}
through the \texttt{userdata} declaration, and the table type
that represents the module.
Now we can use the Typed Lua compiler to check for type errors on
our use of the library:
\begin{verbatim}
    local m = require "md5"
    local x = m.new()
    local y = x:clone()
    local z = m.clone("foo")             -- compile-time error
    print(x:digest() == m.digest(y))
\end{verbatim}

In this example there is a type error because the function \texttt{clone}
expects a value of type \texttt{md5\string_context} instead of a value
of type \texttt{string}.

The description files are the mechanism that we used to include the
typing of the Lua standard library inside Typed Lua.
In Chapter \ref{chap:evaluation} we discuss the issues that we found
while typing the Lua standard library.
